---
title: "Reproducible Research: Peer Assessment 1"
author: "Bruce R"
output: 
  html_document:
    keep_md: true
---

```{r, message=FALSE}
# load packages
library(tidyverse)
library(lubridate)

```


## Loading and preprocessing the data

```{r}
df <- read_csv("activity.csv", show_col_types = FALSE)
# convert date to date format
df$date <- ymd(df$date)
# break apart interval code (hourMinute) and create a Time column.
df <- df %>% 
  mutate(Hour = interval %/% 100,
         Minutes = interval %% 100,
         Time = hms::hms(hours = Hour, minutes = Minutes)
         )

```

Take a peak at data.

```{r}
tail(df)
```


## What is mean total number of steps taken per day?

```{r}
# summarize data by date, ignorning NAs in steps
byDateSummary <- df %>% 
  group_by(date) %>% 
  summarize(stepsPerDay = sum(steps, na.rm = TRUE))

meanStepsPerDay <- as.character(round(mean(byDateSummary$stepsPerDay),2))
medianStepsPerDay <- as.character(median(byDateSummary$stepsPerDay))
```

Histogram for steps each day.

```{r}
ggplot(byDateSummary, aes(x = stepsPerDay)) +
  geom_histogram(binwidth = 1000, fill = "gray", color = "gray30", boundary = 0) +
  labs(title = "Distribution of Number of Steps per Day",
       x = "Number of Steps per Day")
```

The mean number of steps per day is `r meanStepsPerDay`.  
The median number of steps per day is `r medianStepsPerDay`.

## What is the average daily activity pattern?

We take the average steps for each 5-minute interval, removing the NAs from the data. 

```{r}
by5minIntervalSummary <- df %>% 
  filter(!is.na(steps)) %>% 
  group_by(Time) %>% 
  summarize(stepsPerInterval = mean(steps, na.rm = TRUE))

maxTime <- by5minIntervalSummary$Time[which.max(by5minIntervalSummary$stepsPerInterval)]

```

**Note**: I've used the *Time* variable I created instead of *interval* since *interval* 
introduces artificial space between intervals such as 855 and 900.

```{r}
times <- hms(c("0:0:0", "4:0:0", "8:0:0", 
               "12:0:0", "16:0:0", "20:0:0", "24:0:0" ))

ggplot(by5minIntervalSummary, aes(x = Time, y = stepsPerInterval)) +
  geom_line() +
  scale_x_time(breaks = times) +
  labs(title = "Average Number of Steps per 5-minute Inteval",
       x = "5-minute Interval (Time)",
       y = "Average number of steps")

```

The 5-minute interval with the maximum number of steps, on average, starts at the time `r maxTime`.  

## Imputing missing values

```{r}
summary(df)
```

The summary of the data shows that NAs are only in *steps* data and that there
are 2304 NAs in total.

We'll replace these with the average for the corresponding 5-minute interval.

```{r}
# join with average per 5-minute interval
df_complete <- df %>% 
  left_join(by5minIntervalSummary, by = "Time")
# if steps is NA, replace with the average for that time period
df_complete <- df_complete %>% 
  mutate(steps = if_else(is.na(steps), stepsPerInterval, steps))
```

We can see that there no more NAs.

```{r}
summary(df_complete)
```

Summarize by date again with NAs replaced

```{r}
byDateSummary_complete <- df_complete %>% 
  group_by(date) %>% 
  summarize(stepsPerDay = sum(steps, na.rm = TRUE))

meanStepsPerDay_complete <- as.character(round(mean(byDateSummary_complete$stepsPerDay),2))
medianStepsPerDay_complete <- as.character(round(median(byDateSummary_complete$stepsPerDay)))

```


```{r}
ggplot(byDateSummary_complete, aes(x = stepsPerDay)) +
  geom_histogram(binwidth = 1000, fill = "gray", color = "gray30", boundary = 0) +
  labs(title = "Distribution of Number of Steps per Day -- NAs replaced",
       x = "Number of Steps per Day")
```

The mean number of steps per day is `r meanStepsPerDay_complete`.  
The median number of steps per day is `r medianStepsPerDay_complete`.

The mean has increased and many days in the first column are gone.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# add column to signify weekend or weekday -- Sunday = 1, Monday = 2, etc.
df_complete <- df_complete %>% 
  mutate(dayType = if_else(wday(date) == 1 | wday(date) == 7,
                           "Weekend",
                           "Weekday"
                           )
         )

byDayTypeSummary_complete <- df_complete %>% 
  group_by(dayType, Time) %>% 
  summarize(stepsPerInterval = mean(steps, na.rm = TRUE), .groups = 'drop')
```

```{r}
times <- hms(c("0:0:0", "4:0:0", "8:0:0", 
               "12:0:0", "16:0:0", "20:0:0", "24:0:0" ))
ggplot(byDayTypeSummary_complete, aes(x = Time, y = stepsPerInterval)) +
  geom_line() +
  scale_x_time(breaks = times) +
  facet_grid(rows = vars(dayType)) +
  labs(title = "Average Number of Steps per 5-minute Inteval",
       x = "5-minute Interval (Time)",
       y = "Average number of steps")
```
 
 There seems to be a shift to later hours on the weekend and also generally more
 activity in the afternoon on weekends, less in the morning.
